TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-07 manuscript/manuscript.Rmd
Adjust Fig2 name
grammarly edits

---
bibliography: [refs/networks.bib,refs/brain.bib,refs/enrichment.bib,refs/3D_refs.bib]
csl: styles.ref/genomebiology.csl
output:
  pdf_document: default
  word_document:
    reference_docx: styles.doc/Arial_11_single_space_normal_margins.docx
  html_document: default
editor_options: 
  chunk_output_type: console
---

# Coherence of Molecular Mechanisms in Genes Identified Through GWAS In Major Human Disease and Traits

Dozmorov, Mikhail G.^1+^ (mikhail.dozmorov@vcuhealth.org)  
Cresswell, Kellen G.^1+^ (cresswellkg@vcu.edu)  
Reimers, Mark^2^ (reimersm@msu.edu)  
Bacanu, Silviu-Alin^3^ (silviu-alin.bacanu@vcuhealth.org)  
Craver, Carl^4^ (ccraver@wustl.edu)  
Kendler, Kenneth S.^3^ (kenneth.kendler@vcuhealth.org)

^1^ - Department of Biostatistics, Virginia Commonwealth University   
^2^ - Neuroscience Program and Dept. Biomedical Engineering Michigan State  
^3^ - Virginia Institute for Psychiatric and Behavior Genetics and the Department of Psychiatry, Virginia Commonwealth University  
^4^ - Philosophy-Neuroscience-Psychology Program, Washington University in St. Louis  
^+^ - equal contribution

@@

**Background.** Complex phenotypes such as height and intelligence, are thought to be a product of the collective effects of multiple phenotype-associated genes, or, more precisely, interactions among their protein products. High/low degree of interactions is suggestive of coherent/random molecular mechanisms, respectively. Comparing the degree of interactions may help to better predict and understand the coherence of the molecular mechanisms underlying complex phenotypes. However, direct comparison of the degree of interactions is difficult due to differences in size and configuration of phenotype-associated gene networks.

**Methods.** We introduce a measure of network coherence as a ratio of internal vs. external distributions of the degree of interactions. The internal degree distribution is defined by interaction counts within a phenotype-specific gene network, while the external degree distribution counts interactions with other genes in the whole protein-protein interaction (PPI) network. We present a novel method for normalizing the internal degree distributions, making them directly comparable.

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 133 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (the standard of high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric diseases and intelligence-related traits were the least coherent.

**Conclusions.** We present a general-purpose method for estimating and comparing the coherence of phenotype-associated gene networks that accounts for the network size and shape differences. Our results highlight the gaps in of our current knowledge of the genetics and molecular mechanisms of complext phenotypes and suggest priorities for future work.

**Keywords:** GWAS, network, degree, coherence

@@

Genome-wide association studies (GWAS) have significantly advanced our understanding of complex phenotypes by identifying disease- and trait-associated genetic markers and suggesting corresponding genes [@Altshuler:2008aa; @Risch:1996aa]. However, GWAS findings explain only a fraction of heritability [@Eichler:2010aa; @Maher:2008aa; @Manolio:2009aa]. Furthermore, the corresponding genes are often spread across different chromosomes with no known connection to one another, hindering understanding of the molecular mechanisms. These limitations of current generation GWAS might stem from the yet incomplete knowledge of the genetic determinants of complex phenotypes, the potential heterogeneity of complex phenotypes and/or the causes other than genetics [@Botstein:2003aa; @turkheimer2012genome; @Sullivan:2018aa].

Many studies have shown that genetically-driven complex phenotypes are often associated with functionally related genes that are more likely to form networks of interacting protein products [@Hartwell:1999aa; @Krauthammer:2004aa; @Huang:2006aa; @Lage:2007aa; @Cordell:2009aa; @Emily:2009aa; @Feldman:2008aa]. This observation has been verified systematically for a large number of diseases [@Menche:2015aa], thus confirming a fundamental hypothesis of the interactome-based approach to human phenotype, namely that disease- and trait-associated genes tend to form interaction modules [@Goh:2007aa]. Such interaction modules may be viewed as connected subnetworks within the full interactome and may contain all molecular determinants of a certain phenotype. Consequently, methods to identify phenotype-associated networks of functionally-related genes have been developed [@Krauthammer:2004aa; @Cordell:2009aa; @Akula:2011aa].

Network properties, such as connectivity (aka degree distribution) [@Ghiassian:2015aa; @Lee:2008aa; @Barabasi:2004aa; @Feldman:2008aa], can be compared to better understand the relationships among the phenotypes [@Vidal:2011aa; @Barabasi:2011aa; @Wang:2011aa]. Networks formed by phenotype-associated genes, or, more precisely, their protein-protein products, are thought to have high connectivity [@Huang:2006aa; @Krauthammer:2004aa; @Goh:2007aa]. This intuition reflects a well-known "guilt-by-association" principle that genes (or, more precisely, their products) that form an interaction network are more likely to share similar functions [@Farber:2013aa], or co-expression patterns [@Huang:2006aa; @Michalak:2008aa]. Although this view has been criticized [@Gillis:2012aa], several studies consistently observed phenotype-associated genes to be either connected as a single network or to participate in common phenotype-specific subnetworks [@Menche:2015aa; @Lee:2011aa; @Akula:2011aa; @Ideker:2008aa; @Iossifov:2008aa]. Therefore, high connectivity among phenotype-specific genes may indicate coherent molecular mechanisms that could be targeted therapeutically. In terms of graph theory, we are asking whether a phenotype-specific network is a community - a highly connected subgraph relatively well-separated from the rest of the network [@fortunato2010community]. However, the direct comparison of connectivity across phenotype-specific networks is hindered by the fact that different phenotypes are associated with different numbers of genes forming networks of different sizes and configurations.

This study compares the connectivity of networks formed by disease- or trait-associated genes and gene products. We hypothesized that genes associated with (categories of) diseases share common molecular mechanisms and, hence, form connected (i.e., coherent) networks. Therefore, the level of connectivity may inform us about similarities and differences among the diseases. The terms “network connectivity,” “coherence,” or “average degree distribution” are used interchangeably as appropriate. We developed a method for normalizing the internal network connectivity that accounts for differences in network sizes formed by disease-associated genes. We compared the ratio of normalized internal vs. external network connectivity between psychiatric, neurologic, autoimmune, cancer, cardiovascular, metabolic and other diseases and traits. We used KEGG pathways [30] as a reference for a coherent network with common molecular mechanisms, and we used sets of randomly sampled genes as a reference for the absense of coherence. We found that genes associated with metabolic diseases and traits and autoimmune diseases formed coherent networks similar to KEGG pathways. On the contrary, the coherence of networks of psychiatric and neurologic disease-associated genes, as well as intelligence-associated traits, was similar to that of randomly selected genes. Cancer, anthropometric traits, eye-, and cardiovascular diseases and traits formed networks of intermediate connectivity. Our study identifies significant differences between categories of diseases in terms of network properties of the disease-associated genes and highlights the limitations of the current understanding of the molecular determinants of psychiatric disorders.

This study presents a novel measure of network coherence as a ratio of the normalized internal vs. external degree distributions. The internal degree distribution is defined as gene-specific interaction counts within a protein-protein interaction (PPI) network formed by phenotype-associated genes, while the external degree distribution counts interactions with other genes in the whole PPI network. We used KEGG pathways [@Kanehisa:2000aa] as a reference for a paradigmatically high-coherent network [@Lee:2003aa], while sets of randomly sampled genes were used as the reference for the absence of coherence. We hypothesized that the level of coherence may inform us about similarities and differences among phenotypes. Using two PPI databases (STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa]), we compared coherence of 133 phenotypes from GWAScatalog [@Hindorff:2009aa; @MacArthur:2017aa]. Our results show the tendency of complex phenotypes, such as intelligence-related traits, to have low coherence and metabolic traits and diseases to have the most coherence. Our method enables direct comparison of coherence measures and highlight gaps in current understanding of the molecular mechanisms of many phenotypes, e.g., schizophrenia being the lowest coherence disorder in the already low-coherence "Psychiatric disease" category.

@@

## Phenotype-specific networks of protein-protein interactions (PPIs)

We collected 133 phenotype-associated gene lists from the NHGRI-EBI GWAS catalog [@MacArthur:2017aa] (see Methods). These were grouped into seven disease categories (“Autoimmune”, “Cancer”, “Cardiovascular disease”, “Eye disease”, “Metabolic disease”, “Neurologic”, “Psychiatric”) and five trait categories (“Anthropometric trait”, “Cardiovascular trait”, “Eye trait”, “Intelligence”, “Metabolic trait”, Table 1, Supplementary Table 1).

>>>>

---
title: 'TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains'
fontsize: 12pt
mainfont: Times New Roman
csl: styles.ref/genome-research.csl
header-includes:
   - \usepackage{float}
   - \usepackage{setspace}\doublespacing
output:
  word_document:
    reference_docx: styles.doc/Arial_11_single_space_normal_margins.docx
  pdf_document:
    fig_caption: no
  html_document: null
bibliography: [3D_refs.bib,tad_references.bib,difftad_references.bib]
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='figures/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

Kellen G. Cresswell^1,2^ (cresswellkg@vcu.edu), Mikhail G. Dozmorov^1,2\*^ (mikhail.dozmorov@vcuhealth.org)

^1^ Dept. of Biostatistics, Virginia Commonwealth University, Richmond, VA, 23298, USA  
^2^ Equal contribution  
^\*^ To whom correspondence should be addressed: Virginia Commonwealth University, Richmond, VA, 23298, 804-827-2055, mikhail.dozmorov@vcuhealth.org

**Keywords:** Hi-C, chromosome conformation capture, topologically associated domains, TADs, differential analysis, TADCompare

**Running Title:** TADCompare: an R package for differential TAD detection

@@

Recent research using chromatin conformation capture technologies, such as Hi-C, has demonstrated the importance of topologically associated domains (TADs). TADs are considered to be relatively stable three-dimensional (3D) genomic structures orchestrating gene expression regulation; yet, many dynamically reorganize during development or disease, and exhibit cell- and condition-specific differences. Quantification of the dynamic behavior of TADs will help to better understand genome regulation. Methods for comparing TADs between cells and conditions are highly limited. We developed `TADCompare`, a method for differential analysis of TAD boundaries between Hi-C datasets. `TADCompare` is based on a spectral clustering-derived measure called the eigenvector gap, which enables a loci-by-loci comparison of TAD boundary differences between datasets. Using this measure, we introduce methods for identifying differential and consensus TAD boundaries, and tracking TAD boundary changes over time. We further propose a novel framework for the systematic classification of TAD boundary changes. Colocalization- and gene enrichment analysis of different types of TAD boundary changes demonstrated distinct biological functionality associated with them. `TADCompare` is available on https://github.com/dozmorovlab/TADCompare.

### Keywords

@@

Recent research indisputably proves importance of the 3D genome organization in regulating gene expression and other genomic processes [@Schoenfelder:2010ab; @Steensel:2011aa; @Franke:2016aa; @Symmons:2014aa; @Sexton:2015aa; @Li:2012aa; @Papantonis:2013aa; @Laat:2003aa; @Mora:2016aa; @Mifsud:2015aa; @Shavit:2014aa; @Osborne:2004aa; @Schoenfelder:2010aa; @Tanizawa:2010aa]. The 3D genomic structures consists of chromosome territories [@Cremer:2010aa], A/B compartments corresponding to active/repressed chromatin [@Rao:2014aa; @Lieberman-Aiden:2009aa], topologically associated domains (TADs) [@Dixon:2012aa; @Jackson:1998aa; @Ma:1998aa; @Nora:2012aa; @Sexton:2012aa; @Bonev:2017aa; @Dekker:2002aa], smaller sub-TADs [@Phillips-Cremins:2013aa; @Rao:2014aa] and chromatin loops [@Rao:2014aa; @Dowen:2014aa; @Ji:2016aa; @Denker:2016aa]. These structures help to regulate global gene expression [@Schoenfelder:2010ab; @Steensel:2011aa; @Franke:2016aa; @Symmons:2014aa; @Sexton:2015aa; @Li:2012aa; @Papantonis:2013aa; @Laat:2003aa; @Mora:2016aa; @Mifsud:2015aa; @Shavit:2014aa; @Osborne:2004aa; @Schoenfelder:2010aa; @Tanizawa:2010aa]. Consequently, coordinated changes in the 3D structures [@Yaffe:2011aa; @Symmons:2014aa; @Dai:2012aa] determine cell type-specific gene expression and identity [@Dowen:2014aa; @Ji:2016aa; @Phillips-Cremins:2013aa; @Rao:2014aa; @Vietri-Rudan:2015aa; @Jin:2013aa; @Schoenfelder:2010aa; @Dekker:2013aa], guide recombination [@Jhunjhunwala:2009aa], X chromosome inactivation [@Nora:2012aa; @Crane:2015aa]. Many 3D structures are largely invariant between different cell types, and even conserved between mammalian species [@Dixon:2012aa; @Nora:2012aa; @Pope:2014aa; @Naumova:2013aa; @Rao:2014aa; @Vietri-Rudan:2015aa], indicating their high biological importance during genome evolution.

Despite the high level of conservation, recent research uncovered the dynamic nature of the 3D genomic structures, and this plasticity accompanies various biological functions and phenomena [@Yu:2017aa]. In Drosophila, exposure to heat-shock caused local changes in certain TAD boundaries resulting in TAD merging [@Li:2015ac]. Another recent study showed that during motor neuron (MN) differentiation in mammals, TAD, and sub-TAD boundaries in the Hox cluster are not rigid and their plasticity is linked to changes in gene expression during differentiation [@Narendra:2016aa]. The global organization of the 3D genomic structure is found in mitosis [@Nagano:2017aa], <!--zygotic development [@Hug:2017aa; @Flyamer:2017aa], -->earliest stages of mammalian lineage development [@Dixon:2015aa; @Du:2017aa; @Ke:2017aa; @Bonev:2017aa], and somatic cell reprogramming of pluripotent stem cells [@Zhang:2018aa; @Novo:2018aa]. Fusion of TADs [@Flavahan:2016aa; @Fudenberg:2016aa; @Sanborn:2015aa; @Dowen:2014aa; @Guo:2015aa; @Tang:2015aa; @Nora:2012aa], creation or destruction of sub-TADs within existing TAD boundaries [@Taberlay:2016aa; @Lupianez:2016aa], and/or switching TAD states between active and inactive conformations [@Dixon:2012aa; @Lieberman-Aiden:2009aa] has been associated with a variety of phenotypes [@Krijger:2016aa; @Misteli:2010aa; @Spielmann:2018aa], ranging from limb malformation [@Lupianez:2016aa], congenital disorders [@Ibn-Salem:2014aa], to cancer [@Mitelman:2000aa; @Valton:2016aa; @Rickman:2012aa; @Hnisz:2016aa; @Groschel:2014aa; @Corces:2016aa; @Lupianez:2016aa; @Krijger:2016aa; @Flavahan:2016aa; @Barutcu:2015aa]. These observations highlight the importance of studying changes in TADs as a means to understand genomic regulation. However, methods for identifying changes in TAD boundaries remain underdeveloped.

To our knowledge, there are only three methods that can be adapted for detecting changes in TAD boundaries: `localtadsim` [@Sauerwald:2018aa2], `HiCDB` [@Chen:2018aa] and `DiffTAD` [@Zaborowski:2016aa]. Of the three methods, none provide an intuitive, easy to use way of calling differential TADs. Both `localtadsim` and `DiffTAD` are two-step procedures requiring separately defined TADs and comparing them using a command-line utility. `HiCDB` has a built-in TAD caller but does not allow for comparisons of chromosome-specific contact matrices. All three methods require highly specific data types and file names to be able to run. The lack of usability is compounded with issues such as a lack of upkeep, slow runtimes, and lack of statistical rigor (Supplementary methods).

As costs of Hi-C data continue to drop, several studies started to investigate the dynamics of 3D changes over time. The most notable applications include cell differentiation studies [@Bonev:2017aa], embryonic development [@Du:2017aa;@Hug:2017aa;@Ke:2017aa], cancer progression [@Zhou:2019aa]. Typically, TAD boundary changes over time are quantified by overlap [@Hug:2017aa;@Du:2017aa] and classified into distinct patterns [@Zhou:2019aa]. However, general-purpose methods for systematic analysis of TAD boundary changes over time do not exist. Furthermore, the number of replicates for a given experiment continues to rise, requiring methods for defining and comparing TAD boundaries across replicates of Hi-C data. <!--Previous research has shown inconsistencies between both technological and biological replicates [@Dixon:2012aa;@Rao:2014aa;@Sauerwald:2018aa2]. As a result, researchers must take this variation into account and make full use of the data. -->Traditionally, two approaches have been developed to identify TAD boundaries across replicates. The first is to call TADs on individual replicates and aggregate them (e.g., `Arrowhead` [@Rao:2014aa]). The second approach involves combining all replicates into a consensus contact matrix <!--using utilities such as `Juicer` [@Durand:2016aa] -->and then calling TADs (`Arboretum` [@Fotuhi-Siahpirani:2016aa]). [??? TBD - are the tools mentioned to the corresponding methods?] However, it is focused on identifying TADs across cell-lines and species as opposed to across replicates. Methods for detecting consensus TADs across Hi-C datasets remain underdeveloped.

We developed `TADCompare`, an R package aimed at providing a fast, accurate, user-friendly and well-documented approach to differential TAD analysis. We introduce a method based on the boundary score statistic [@Cresswell:2019aa] and use it to identify five types of TAD boundary changes. The method is extended to allow for calling consensus TAD boundaries and comparing them between groups of Hi-C replicates. We further demonstrate how the TAD boundary score statistic may be used to analyze TAD dynamics over the time course. For both differential TAD boundary detection and time course analysis, we provide novel terminology for the classification of TAD boundary changes. We demonstrated the robustness of `TADCompare` using simulated data with pre-defined TADs [@Forcato:2017aa] and its ability to reveal distinct biological roles of different TAD boundary changes. In summary, `TADCompare` provides an all-in-one pipeline from TAD calling to differential boundary detection, including time course, that supports replicated Hi-C data analysis. The output is formatted in a commonly used BED format that allows for flexible downstream analyses and visualization. The `TADCompare` R package is freely available on GitHub (https://github.com/dozmorovlab/TADCompare).

### Representation of Hi-C data as a graph

For a given Hi-C experiment, Hi-C data is represented by a chromosome-specific contact matrix $C$ of non-overlapping regions (aka bins) of size $r$ (resolution of the data). Each entry $C_{ij}$ corresponds to the number of contacts between region $i$ and region $j$. Previous work has shown that this contact matrix is essentially an analog of the adjacency matrix found in graph theory and Hi-C data can be thought of as a naturally occurring graph where edges are contacts and vertices are genomic regions [@Boulos:2013aa; @wang2017network; @Wang:2013:TPC:2506583.2506633;@Cresswell:2019aa]. The graph representation of Hi-C data is the foundation of our method and allows us to use a graph-clustering based approach to identify and analyze TADs.

### Calculating the graph spectrum

The first step of our method is to calculate the graph spectrum, defined as the eigenvectors of the Laplacian of an adjacency matrix. Using the interpretation of the contact matrix as a naturally occurring adjacency matrix, we calculate the Laplacian directly from the contact data. Briefly, the graph spectrum for a given contact matrix is calculated as follows:

1. Calculate the normalized Laplacian $\bar{L}$:

$$
\bar{L} = D^{-\frac{1}{2}}CD^{-\frac{1}{2}}
$$

where $D = diag(\textbf{1}^{T}C)$, where $\textbf{1}$ is a column vector of size $C$ where each entry is 1. $D$ can be thought of as a vector containing the sum of the degrees for a given node.

2. Perform an eigendecompositon of the Laplacian:

$$
\bar{L}v = \lambda v
$$

In practice, we calculate the first two eigenvectors with the largest absolute values of eigenvalues and organize them into a matrix $\bar{V}$ with dimensions $i \times 2$, where $i$ is the number of regions in the contact matrix. $\bar{V}$ is referred to as the graph spectrum of the contact matrix.

### Eigenvector gap as a measure of pattern change

We can think of each row of the matrix $\bar{V}$ as a quantification of the pattern of contacts in each region of the contact matrix. Previous work [@Cresswell:2019aa] has demonstrated that by taking the Euclidean distance between row $V_{i.}$ and its neighboring row $V_{(i+1).}$, one can measure the similarity in the pattern of contacts between region $i$ and region $i+1$ of the chromosome, termed "eigenvector gap". A TAD boundary manifests itself as a sudden break in the pattern of contacts. This pattern is reflected in the eigenvector gap by a spike in gap size followed by and preceded by smaller gaps (Figure 1). The eigenvector gap quantifies the degree of this break, acting as a proxy for TAD boundary likelihood. To calculate the eigenvector gaps, we perform the following procedure:

1. Normalize columns of $\bar{V}$ to sum to 1:

$$
\begin{aligned}
\hat{V_{ij}} = \frac{\bar{V_{ij}}}{ \Vert{\bar{V_{.j}\Vert}}}
\end{aligned}
$$

where the subscript $.j$ corresponds to column $j$.

2. Normalize $\hat{V}$ and project onto a unit circle:

$$
\tilde{Z} = Diag(diag^{-\frac{1}{2}}(\hat{V_{i.}}\hat{V_{i.}}^{T}))\hat{V_{i.}}
$$
<!--$\tilde{Z}$ is the projection of $\bar{V}$ onto a unit circle such that all values are bounded by [-1,1].-->

3. Calculate the distance between neighboring regions (rows $i$ and $i-1$ of $\tilde{Z}$) and store in a vector $D_{i}$:

$$
D_{i}=\sqrt{(\tilde{Z}_{i1}-\tilde{Z}_{(i-1)1})^2 + (\tilde{Z}_{i2}-\tilde{Z}_{(i-1)2})^2}
$$

We refer to $D$ as the vector where each entry $D_i$ is referred to as an eigenvector gap. Formally, an eigenvector gap is the Euclidean distance between each successive row of the first two eigenvectors. In practical terms, the eigenvector gap for a given locus is a measure of how likely that locus is a TAD boundary.

To maintain the association of each entry of the vector with its corresponding matrix region, a placeholder is used in the first entry of the vector. This is necessary because we cannot calculate an eigenvector gap for the first entry of the contact matrix due to a lack of a left-bound neighbor. In mathematical terms, this means that for a matrix of size $n$ the total number of eigenvector gaps is $n-1$.

<!--
Graph spectrum calculation scales poorly for large matrices, having a quadratic complexity [@Wu:2018aa]. By using a windowed approach, the complexity becomes linear, relying mainly on the window size to determine the speed of the algorithm. The window starts at the beginning of the contact matrix. After calculating the eigenvector gaps within the window, the window moves forward $w-1$ bins across the diagonal where $w$ is the size of the window. We move $w-1$ spaces because for a given window it is impossible to calculate the eigenvector gap of the last bin. Moving $w-1$ allows us to include that last bin in the next window and calculate this value. We continue this process until the end of the contact matrix is reached.  
-->

### Converting eigenvector gaps to boundary scores

We showed that the distribution of eigenvector gaps can be approximated by a log-normal distribution (Supplementary Figure 1, Supplementary Material). The log-normality allows us to convert the eigenvector gap values into boundary scores:

$$
B_{i} = \frac{ (ln(D_i) - \mu)} {\sigma^{2}}
$$

where $ln(D) \sim N(\mu, \sigma^{2})$ where $\mu$ and $\sigma^{2}$ are the mean and variance of the distribution of the natural log of the eigenvector gaps, respectively, and $B$ is a vector of boundary scores with a N(0,1) distribution. In practice, this value is simply the Z-score for the natural log of eigenvector gaps.

### Sliding window eigenvector gap calculation

Frequency of interactions decays following power law as the distance between the interacting regions in a linear genome increases [@Lajoie:2015aa]. This decay leads to noisy and non-informative interactions farther off-diagonal of the contact matrix. To alleviate the effect of noisy distant interactions, we perform spectral decomposition within a fixed-size window that moves along the diagonal of the matrix. For instance, a window size of 15 bins (default setting, Supplementary Figure 2, Supplementary Material) means that only values within 15 bins of the diagonal will be used to calculate the eigenvector gap. The sliding window approach improves the performance of eigenvector gap calculation [@Cresswell:2019aa]. Additionally, it provides for faster calculations, operating on many small matrices instead of one large matrix.

<!-- ### Measuring contact matrix similarity using boundary scores -->

<!-- Oftentimes, it is of interest to measure the similarity of entire contact matrices. This is useful for comparing technical or biological replicates and assessing the consistency of Hi-C sequencing. Calculating similarity is complicated by the need to capture the topological similarity of contact matrices and certain issues such as distance decay and noise. Previous methods have attempted to quantify this similarity, such as Quasar-Rep [@Sauria:2015aa], Hi-C-Spector[@Yardimci:2017aa], Hi-CRep [@Yang:2017aa] and GenomeDISCO [@Ursu:2018aa]. Both the Quasar-Rep and approach takes two contact matrices, transforms them by stratifying based on distance, and then takes a Pearson correlation. GenomeDISCO and Hi-C-Spector are graph-based methods, with GenomeDISCO measuring similarity based on the distance between graph-diffusion smoothed contact matrices and Hi-C-Spector measuring the distance between eigenvectors of the Laplacians of each contact matrix. We propose a method based on cross-correlation of boundaries to calculate similarity. -->

<!-- Given two contact matrices $P$ and $R$ with boundary scores of $B_P$ and $B_R$ having means of $\mu_{P}$ and $\mu_{R}$ respectively, we can calculate matrix similarity using a modified form of cross-correlation that takes into account small shifts in TAD boundaries. For this example, we first define the basic cross-correlation of two sets of boundary scores with lag $l$: -->

<!-- $$ -->
<!-- S = \frac{ \sum[(B_{P}(i)-\mu_{P})*(B_{R}(i-l)-\mu_{R})] -->

<!-- }{\sqrt{\sum(B_{P}(i)-\mu_{P})^{2}}\sqrt{\sum(B_{R}(i-l)-\mu_{R})^{2}}} -->
<!-- $$ -->

<!-- where $i$ is the corresponding vector entry and $l$ is the lag. Using this framework, we incorporate the fact that shifts of less than 5 are biologically insignificant by replacing all instances of $B_{R}(i-l)$ with $B_{min}(i) = min(|(B_{P}(i)-\mu_{P})-(B_{R}(i-l)-\mu_{R})|)$ for l = 1,...,5. The resulting formula is the following: -->

<!-- $$ -->
<!-- S_{adj} = \frac{ \sum_{i}[(B_{P}(i)-\mu_{P})*(B_{min}(i)-\mu_{R})] -->

<!-- }{\sqrt{\sum_{i}(B_{P}(i)-\mu_{P})^{2}}\sqrt{\sum_{i}(B_{min}(i)-\mu_{R})^{2}}} -->
<!-- $$ -->

<!-- $S_{adj}$ gives us a measure of similarity between contact matrices that gives equal weighting to lagged and non-lagged TAD boundaries, finding the region of maximum similarity. This score effectively provides a measure of similarity between contact matrices purely in terms of TAD similarity. The adjustment allows us to treat shifted boundaries the same as non-shifted boundaries and more accurately capture contact matrix similarity.   -->

### Differential analysis using boundary scores

To define the differences between two contact matrices, $P$ and $R$, we compare their eigenvector gaps $D_P$ and $D_R$, respectively. Given that $ln(D_P) \sim N(\mu_{P}, \sigma_{P}^{2})$ and $ln(D_R) \sim N(\mu_{R}, \sigma_{R}^{2})$, it follows that $ln(D_P) - ln(D_R) \sim N(\mu_{P}-\mu_{R}, \sigma_{P}^{2}+\sigma_{R}^{2})$. These results allow us to calculate a vector of differential boundary scores:

$$
DB_{i} = \frac{ (ln(D_{Pi}) - ln(D_{Ri})) - (\mu_{P}-\mu_{R}) } {\sigma_{P}^{2}+\sigma_{R}^{2}}
$$

or more simply,

$$
DB_{i} = \frac{\sigma_{P}^{2}B_{P}-\sigma_{R}^{2}B_{R}}{\sigma_{P}^{2}+\sigma_{R}^{2}}
$$

where $B_{P}$ and $B_{R}$ are the boundary scores for the $P$ and $R$ matrices, respectively. This score can be thought of as the difference in TAD boundary likelihood for a given locus in two data sets. Due to the aforementioned normality of the difference in log eigenvector gaps, $DB_i$ can be thought of as a simple z-score where $DB \sim N(0,1)$.

### Time course boundary changes

Boundary scores provide a convenient method for modeling the change of TAD boundaries over time. For a given TAD boundary, or, any region of the genome, we can monitor the trajectory of the boundary score. Over time, we can define boundary score changes based on their deviation from a baseline level (typically, the first time point). It is expected that these scores will be relatively constant over time except in regions where a boundary appears or disappears. The trend across time points can be recorded, and the pattern of change classified accordingly. Our implementation of time course boundary analysis allows for the usage of multiple replicates for a given time point. Briefly, at each region of the genome, the consensus boundary score is calculated, defined as the median of consensus scores across all replicates, and is then used to identify boundaries.

<!--The overall trend can be calculated by taking the full set of TAD boundaries across every time point and calculating the average boundary score for each individual point. If this score is increasing, then it suggests the appearance of TAD boundaries over time. Conversely, if it is decreasing, then there is a disappearance of TAD boundaries. We can also incorporate the consensus boundary score framework into boundary score calculations over time. Given a set of replicates at multiple different time points, we can call consensus boundary scores at for each time points and aggregate them. We can also quantify the change in TAD strength for the entire contact matrix over time by modeling the trend of mean boundary scores across each of the time point. A positive trend indicates an increase in TAD strength or abundance and a decrease indicates a drop.-->

<!--
### Simulated data for differential TAD boundary detection

To simulate differential TAD boundaries, we used 25 simulated contact matrices with manually annotated TADs downloaded from Hi-CToolsCompare [@Forcato:2017aa]. These matrices were generated at five levels of noise (4%, 8%, 12%, 16%, and 20%) with five simulated matrices per noise level. Noise is generated by randomly selecting entries in the contact matrix and adding two contacts. The percentage value corresponds to the percentage of matrix entries with constant noise added.

Within each noise level, we take each possible combination of contact matrices and calculate differential TADs between them. It is expected that an accurate differential boundary detection method will find all manually annotated TADs which differ between each combination of simulated matrices. For each combination of simulated matrices, we define true differential TAD boundaries as the set of non-shared boundaries between the matrices. To assess the ability of our method to discriminate between true differential TADs and non-differential TADs, we use the Matthews correlation coefficient. [??? TBD, rationale for MCC]
-->
<!-- ### Primary and replicate experimental data for differential TAD boundary detection -->

<!-- Primary and replicate data from [@Rao:2014aa] was used to test the percentage of differential TAD boundaries between primary and replicate data. TAD boundaries were called using the `SpectralTAD` TAD caller [@Cresswell:2019aa]. The raw percentage of shared boundaries at the first level of the hierarchy were calculated between the primary and replicate. To test whether these boundaries were truly differential, we then ran `TADCompare` and calculated the percentage of non-shared boundaries which were also detected by TADCompare.  -->

<!-- We hypothesized that non-shared boundaries between primary and replicates are largely the result of differences in the TAD hierarchy. If true, it is expected that TAD boundaries which are not shared between primary and replicate and are not detected by `TADCompare` will actually appear somewhere in both the primary and replicate's hierarchy of TADs. [??? Shall we forgo hierarchy?] -->
<!--
### Handling of gaps

Gaps refer to loci with either complete absence of loci or levels of sparsity which make analysis impossible. For our analysis, we defined gaps as any region with less than 80\% of its contacts greater than zero. Since we are not particularly interested in long-range contacts, this percentage is calculated based on regions within our sliding window. These gaps can introduce instability in the algorithm and lack important information. To counter this, we simply remove these regions before the analysis. This is done for both contact matrices such that, if one contact matrix contains a gap at a given locus and the other does not, we remove it from both. This allows us to still make a one-to-one comparison of regions. 
-->

### Data sources

All simulated data were downloaded from `HiCToolsCompare` repository [@Forcato:2017aa]. In total, we used 25 simulated matrices with varying levels of noise. For sparsity and downsampling analysis matrices were manually created based on matrices from `HiCToolsCompare` matrices with the minimum noise level (see [@Cresswell:2019aa] for methods description). Data for comparisons across cell lines, replicates, and tissues were taken from [@Schmitt:2016aa], generated at 40kb resolution (Supplementary Table 1). Time course data was taken from [@Rao:2017aa], HCT-116 human colon cancer cell-line at four time points after auxin-treatment withdrawal (20, 40, 60, 180 minutes). Contact matrices were generated at 25kb, 50kb, and 100kb using the `straw` tool from `Juicer` [@Durand:2016aa]. Chromatin state data were taken from chromHMM [@Ernst:2010aa]. Histone modifications and transcription factor binding sites were downloaded from the Encyclopedia of DNA Elements (ENCODE) [@Davis:2018aa] (Supplementary Table 2).

### Gene enrichment testing

All gene enrichment testing was performed using the `rGREAT` (Version 2.0) R package. Briefly, we detect genes within 5kb upstream and 1kb downstream of each type of TAD boundary changes, similar to work of others [@Chen:2018aa]. For each Gene Ontology (GO) and pathway (MSigDb [??? Describe version, add ref]), a hypergeometric test is then performed to determine over-representation of TAD boundary-associated genes. For all figures, we report results for GO Biological Processes. Results for GO Molecular Function, GO Cellular Component, MSigDB and PANTHER pathways are reported in tables.

### Colocalization enrichment testing

A permutation test was used to quantify the enrichment of colocalization of TAD boundaries of interest with genomic annotations. Briefly, we flank each type of TAD boundary changes (differential or time course) by 50kb on each side and calculate the mean number of genomic annotations across those regions (observed enrichment). Next, we generate two sets of bins, one the size of the TAD boundaries which we are testing (considering the flanking) and another the size of all other bins. The difference in the mean number of genomic annotations colocalized with TAD boundaries of interest was calculated for each set (expected enrichment). We repeat this procedure 10,000 times. We calculate the permutation p-value by taking the number of times the expected enrichment was greater than the observed enrichment, and dividing by 10000. $\alpha = 0.05$ was set to assess statistical significance.

@@

### A modified spectral clustering approach is better suited for TAD boundary detection than other approaches

Our previous work on TAD detection using spectral clustering, implemented as a `SpectralTAD` R package [@Cresswell:2019aa], introduced the concept of the boundary score statistic, adapted here for differential boundary detection. Briefly, the boundary score is calculated for each bin by sliding a window across the diagonal of the contact matrix, calculating the eigenvectors of the Laplacian matrix, finding the distance between consecutive eigenvectors (eigenvector gap) and converting them into Z-scores (boundary score, see Methods). The boundary score is a continuous measure of TAD boundary likelihood.
